---
title: "Clinical Trials Compliance Study"
output: github_document
always_allow_html: true
---

## Highly-Likely Applicable Clinical Trials Analysis

Table 1 of the [original article](https://www.nejm.org/doi/full/10.1056/NEJMsa1409364#sec-2) groups the 13,327 HLACTs in three categories: all trials, those that reported results in the twelve-month reporting window, and those that reported results in the five-year reporting window.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(knitr.kable.NA = '')
```

### All trials categorized and ranked by purpose

The total set of HLACTs found in the Clinical Trials database (count: 14467) demonstrate a similar spread of `primary_purpose` as the original Anderson 2015 dataset of 13327 HLACTs.

```{r echo=FALSE, message=FALSE}
library(tidyverse)
library(kableExtra)
library(gtsummary)
library(tinytex)

init <- arrow::read_parquet("../brick/analysis-20130927/ctgov-studies-hlact.parquet")

twelve <- init %>%  
  mutate(results_received_12 = months_to_report_results <= 13) %>%
  filter(results_received_12 == TRUE)

five <- init %>%
  mutate(results_received_5 = months_to_report_results <= 65) %>%
  filter(results_received_5 == TRUE)

# all results
purpose <- group_by(init, primary_purpose) %>% count(primary_purpose, name = "primary_purpose_total")
purpose_total <- sum(purpose$primary_purpose_total)
purpose <- purpose %>%
  mutate(primary_purpose_percentage = round(primary_purpose_total / purpose_total * 100, digits=2)) %>%
  filter(primary_purpose == "TREATMENT" || primary_purpose == "PREVENTION")

twelve_purpose <- twelve %>%
  group_by(primary_purpose) %>%
  count(primary_purpose, name = "primary_purpose_total")
twelve_purpose_total <- sum(twelve_purpose$primary_purpose_total)
with_sum <- twelve_purpose %>%
  mutate(primary_purpose_percentage = round(primary_purpose_total / twelve_purpose_total * 100, digits=2)) %>%
  filter(primary_purpose == "TREATMENT" || primary_purpose == "PREVENTION")
  
five_purpose <- five %>%
  group_by(primary_purpose) %>%
  count(primary_purpose, name = "primary_purpose_total")
five_purpose_total <- sum(five_purpose$primary_purpose_total)
five_with_sum <- five_purpose %>%
  mutate(primary_purpose_percentage = round(primary_purpose_total / five_purpose_total * 100, digits=2)) %>%
  filter(primary_purpose == "TREATMENT" || primary_purpose == "PREVENTION")


ordered <- purpose[order(purpose$primary_purpose_percentage, decreasing=TRUE),]
```
#### Trial purpose
```{r echo=FALSE, message=FALSE}
test_joined <- ordered %>%
  left_join(with_sum, by="primary_purpose", suffix = c("_all", "_one_year")) %>%
  left_join(five_with_sum, by="primary_purpose", suffix = c(" (all)", " (five_year)")) %>%
  select(!contains("total")) %>%
  rename(All = primary_purpose_percentage_all, "Twelve months" = primary_purpose_percentage_one_year, "Five years" = primary_purpose_percentage)

test_joined %>%
  knitr::kable(
    format = "markdown",
    align = "l",
    booktabs = TRUE,
    longtable = TRUE,
    linesep = "",
  )  %>%
  kableExtra::kable_styling(
      position = "left",
      latex_options = c("striped", "repeat_header"),
      stripe_color = "gray!15"
    )
```

### Intervention type

The main inclusion criterion for the original study was that the trial have an interventional study design if studying a drug. The breakdown of intervention type for the current HLACT dataset is as follows:

```{r echo=FALSE, message=FALSE}

i_type <- group_by(init, intervention_type) %>% count(intervention_type, name = "intervention_type_total")
i_type_total <- sum(i_type$intervention_type_total)
i_type <- i_type %>%
  mutate(intervention_type_percentage = round(intervention_type_total / i_type_total * 100, digits=2)) %>%
  filter("DRUG" %in% intervention_type || "DEVICE" %in% intervention_type || "BIOLOGICAL" %in% intervention_type)

twelve_i_type <- group_by(twelve, intervention_type) %>% count(intervention_type, name = "intervention_type_total")
t_total <- sum(twelve_i_type$intervention_type_total)
twelve_i_type <- twelve_i_type %>%
  mutate(intervention_type_percentage = round(intervention_type_total / t_total * 100, digits=2)) %>%
  filter("DRUG" %in% intervention_type || "DEVICE" %in% intervention_type || "BIOLOGICAL" %in% intervention_type)

five_i_type <- group_by(five, intervention_type) %>% count(intervention_type, name = "intervention_type_total")
ft_total <- sum(five_i_type$intervention_type_total)
five_i_type <- five_i_type %>%
  mutate(intervention_type_percentage = round(intervention_type_total / ft_total * 100, digits=2)) %>%
  filter("DRUG" %in% intervention_type || "DEVICE" %in% intervention_type || "BIOLOGICAL" %in% intervention_type)

intervention_ordered <- i_type[order(i_type$intervention_type_percentage, decreasing=TRUE),]

i_type_joined <- left_join(intervention_ordered, twelve_i_type, by="intervention_type", suffix = c("_all", "_one_year")) %>%
  left_join(five_i_type, by="intervention_type") %>%
  select(!contains("total")) %>%
  rename(All = "intervention_type_percentage_all", "One year" = "intervention_type_percentage_one_year", "Five years" = "intervention_type_percentage")

i_type_joined %>%
  knitr::kable(
    format = "markdown",
    align = "l",
    booktabs = TRUE,
    longtable = TRUE,
    linesep = "",
  )  %>%
  kableExtra::kable_styling(
      position = "left",
      latex_options = c("striped", "repeat_header"),
      stripe_color = "gray!15"
    )
```

### Trial phases

Trials were considered if they were not in Phase 0, or Early Phase 1.

```{r echo=FALSE, message=FALSE}

phase_df <- group_by(init, phase) %>% count(phase, name = "phase_total")
p_total <- sum(phase_df$phase_total)
phase_df <- phase_df %>%
  mutate(phase_percentage = phase_total / p_total * 100) %>%
  filter(phase == "PHASE2" || 
         phase == "PHASE3" || 
         phase == "PHASE4" ||
         phase == "PHASE2; PHASE3" ||
         phase == "PHASE1; PHASE2" ||
         phase == "NA")

twelve_phase <- group_by(twelve, phase) %>% count(phase, name= "phase_total")
tp_total <- sum(twelve_phase$phase_total)
twelve_phase <- twelve_phase %>%
  mutate(phase_percentage = round(phase_total / tp_total * 100, digits=2)) %>%
  filter(phase == "PHASE2" || 
         phase == "PHASE3" || 
         phase == "PHASE4" ||
         phase == "PHASE2; PHASE3" ||
         phase == "PHASE1; PHASE2" ||
         phase == "NA")

five_phase <- group_by(five, phase) %>% count(phase, name= "phase_total")
fp_total <- sum(five_phase$phase_total)
five_phase <- five_phase %>%
  mutate(phase_percentage = round(phase_total / fp_total * 100, digits=2)) %>%
  filter(phase == "PHASE2" || 
         phase == "PHASE3" || 
         phase == "PHASE4" ||
         phase == "PHASE2; PHASE3" ||
         phase == "PHASE1; PHASE2" ||
         phase == "NA")

phase_ordered <- c("PHASE1; PHASE2", "PHASE2", "PHASE2; PHASE3",
                     "PHASE3", "PHASE4", "NA")

phase_df <- phase_df %>%
  mutate(phase = factor(phase, levels = phase_ordered)) %>%
  arrange(phase)

phase_joined <- phase_df %>%
  left_join(twelve_phase, by = "phase", suffix = c("_all", "_one_year")) %>%
  left_join(five_phase, by = "phase") %>%
  select(!contains("total")) %>%
  rename(All = "phase_percentage_all", "One year" = "phase_percentage_one_year", "Five years" = "phase_percentage")

phase_joined %>%
  knitr::kable(
    format = "markdown",
    align = "l",
    booktabs = TRUE,
    longtable = TRUE,
    linesep = "",
  )  %>%
  kableExtra::kable_styling(
      position = "left",
      latex_options = c("striped", "repeat_header"),
      stripe_color = "gray!15"
    )
```

### Trial site

Many trials in the Clinical Trials dataset were conducted at multiple sites, sometimes across several countries and continents. A major factor in the original analysis was the presence of at least one US site in the trial.

```{r echo=FALSE, message=FALSE}

trial_sites <- group_by(init, has_us_facility) %>% count(has_us_facility, name = "has_us_site")
total <- sum(trial_sites$has_us_site)
trial_sites <- trial_sites %>%
  mutate(us_site_percentage = round(has_us_site / total * 100, digits=2))

twelve_sites <- group_by(twelve, has_us_facility) %>% count(has_us_facility, name = "has_us_site")
t_site_total <- sum(twelve_sites$has_us_site)
twelve_sites <- twelve_sites %>%
  mutate(us_site_percentage = round(has_us_site / t_site_total * 100, digits =2))

five_sites <- group_by(five, has_us_facility) %>% count(has_us_facility, name = "has_us_site")
f_site_total <- sum(five_sites$has_us_site)
five_sites <- five_sites %>%
  mutate(us_site_percentage = round(has_us_site / f_site_total * 100, digits =2))

site_ordered <- trial_sites[order(trial_sites$us_site_percentage, decreasing=TRUE),] %>%
  left_join(twelve_sites, by="has_us_facility", suffix = c("_all", "_one_year")) %>%
  left_join(five_sites, by="has_us_facility") %>%
  select(!contains("has_us_site")) %>%
  rename(All = "us_site_percentage_all", "One year" = "us_site_percentage_one_year", "Five years" = "us_site_percentage")

site_ordered %>%
  knitr::kable(
    format = "markdown",
    align = "l",
    booktabs = TRUE,
    longtable = TRUE,
    linesep = "",
  )  %>%
  kableExtra::kable_styling(
      position = "left",
      latex_options = c("striped", "repeat_header"),
      stripe_color = "gray!15"
    )
```

### Funding source

The available data enumerates three sponsor types for clinical trials: National Institutes of Health (NIH), industry, and other government or academic institution. In Anderson, 2015, industry-led trials showed much higher compliance with reporting requirements during the initial twelve-month period than government- or academic-run trials. Over the five-year period, these gaps closed, but total reporting rates were still less than 50%.

```{r echo=FALSE, message=FALSE}

source <- group_by(init, funding_source) %>% count(funding_source, name = "source_totals")
s_totals <- sum(source$source_totals)
source <- source %>%
  mutate(source_percentage = round(source_totals / s_totals * 100, digits=2)) %>%
  filter(funding_source == "OTHER" || funding_source == "INDUSTRY" || funding_source == "NIH")

twelve_source <- group_by(twelve, funding_source) %>% count(funding_source, name ="source_totals")
t_source_totals <- sum(twelve_source$source_totals)
twelve_source <- twelve_source %>%
  mutate(source_percentage = round(source_totals / t_source_totals * 100, digits=2)) %>%
  filter(funding_source == "OTHER" || funding_source == "INDUSTRY" || funding_source == "NIH")

five_source <- group_by(five, funding_source) %>% count(funding_source, name ="source_totals")
f_source_totals <- sum(five_source$source_totals)
five_source <- five_source %>%
  mutate(source_percentage = round(source_totals / f_source_totals * 100, digits=2)) %>%
  filter(funding_source == "OTHER" || funding_source == "INDUSTRY" || funding_source == "NIH")

ordered_sources <- source[order(source$source_percentage, decreasing=TRUE),] %>%
  left_join(twelve_source, by = "funding_source", suffix = c("_all", "_one_year")) %>%
  left_join(five_source, by = "funding_source") %>%
  select(!contains("totals")) %>%
  rename(All = "source_percentage_all", "One year" = "source_percentage_one_year", "Five years" = "source_percentage")

ordered_sources %>%
  knitr::kable(
    format = "markdown",
    align = "l",
    booktabs = TRUE,
    longtable = TRUE,
    linesep = "",
  )  %>%
  kableExtra::kable_styling(
      position = "left",
      latex_options = c("striped", "repeat_header"),
      stripe_color = "gray!15"
    )
```

### Categorized by completion date

```{r echo=FALSE, message=FALSE}

dates <- init %>%
  mutate(primary_completion_date = as.Date(ym(primary_completion_date)),
         verification_date = as.Date(case_when(
           !is.na(verification_date) & between(ym(verification_date), ym("2008-01"), ym("2012-09")) ~ ym(verification_date),
           TRUE ~ NA_Date_
         )),
         completion_date = as.Date(ym(completion_date)))

twelve_dates <- twelve %>%
  mutate(primary_completion_date = as.Date(ym(primary_completion_date)),
         verification_date = as.Date(case_when(
           !is.na(verification_date) & between(ym(verification_date), ym("2008-01"), ym("2012-09")) ~ ym(verification_date),
           TRUE ~ NA_Date_
         )),
         completion_date = as.Date(ym(completion_date)))

twelve_years <- twelve_dates %>%
  mutate(primary_completion_date_imputed = coalesce(
    primary_completion_date, verification_date, completion_date
  )) %>% 
  mutate(primary_completion_date_imputed = year(primary_completion_date_imputed)) %>%
  group_by(primary_completion_date_imputed) %>% count(primary_completion_date_imputed, name = "count")

five_dates <- five %>%
  mutate(primary_completion_date = as.Date(ym(primary_completion_date)),
         verification_date = as.Date(case_when(
           !is.na(verification_date) & between(ym(verification_date), ym("2008-01"), ym("2012-09")) ~ ym(verification_date),
           TRUE ~ NA_Date_
         )),
         completion_date = as.Date(ym(completion_date)))

five_years <- five_dates %>%
  mutate(primary_completion_date_imputed = coalesce(
    primary_completion_date, verification_date, completion_date
  )) %>% 
  mutate(primary_completion_date_imputed = year(primary_completion_date_imputed)) %>%
  group_by(primary_completion_date_imputed) %>% count(primary_completion_date_imputed, name = "count")

years <- dates %>%
  mutate(primary_completion_date_imputed = coalesce(
    primary_completion_date, verification_date, completion_date
  )) %>% 
  mutate(primary_completion_date_imputed = year(primary_completion_date_imputed)) %>%
  group_by(primary_completion_date_imputed) %>% count(primary_completion_date_imputed, name = "count")

totals <- sum(years$count)
twelve_totals <- sum(twelve_years$count)
five_totals <- sum(five_years$count)
years <- years %>%
  mutate(years_percentage = round(count / totals * 100, digits=2))
twelve_years <- twelve_years %>%
  mutate(years_percentage = round(count/ twelve_totals * 100, digits =2))
five_years <- five_years %>%
  mutate(years_percentage = round(count/ five_totals * 100, digits = 2))

years_joined <- years %>% 
  left_join(twelve_years, by = "primary_completion_date_imputed", suffix = c("_all", "_one_year")) %>%
  left_join(five_years, by = "primary_completion_date_imputed") %>%
  select(!contains("count")) %>%
  rename(All = "years_percentage_all", "One year" = "years_percentage_one_year", "Five years" = "years_percentage")

years_joined %>%
  knitr::kable(
    format = "markdown",
    align = "l",
    booktabs = TRUE,
    longtable = TRUE,
    linesep = "",
  )  %>%
  kableExtra::kable_styling(
      position = "left",
      latex_options = c("striped", "repeat_header"),
      stripe_color = "gray!15"
    )
```
